Supplementary Materials

### 2025-07-07 point-by-point-response.qmd

> - Discuss if there are methods to reduce missing data for the definitive phase.

We have added a suggestion to the discussion to improve the completeness of the data extracted from the medical records.

> - Comment on the ethics of the full trial. Is it ethical to have sites continue with usual care after seeing the pilot study results?

We have added a sentence on the ethics of the full trial.

> - Consider a stepped wedge design where all clusters start with standard of care and then transition into an intervention.

Thank you for this suggestion and we agree that a stepped wedge design would be a good approach for the full scale trial.

> - Use the CONSORT statement extension for pilot and feasibility trials (https://www.equator-network.org/reporting-guidelines/consort-2010-statement-extension-to-randomised-pilot-and-feasibility-trials/) and the CONSORT statement extension for cluster randomised trials (https://www.equator-network.org/reporting-guidelines/consort-cluster/).

We have attempted to align our reporting with these guidelines.

@@

> This is a well written paper, with transparent and thorough methodological reporting, and a reliable statistical approach. The purpose of the analysis was to inform a decision on the feasability of a full trial, and as such the results are interpreted with appropriate caution.
>
> I have only minor comments about certain study design aspects which I found unclear - particularly the observed/extracted data comparison.

Thank you for reviewing the manuscript!

> Methods:
> 1. Intervention – how long was the training programme, and who was it delivered by? Was the training performed in stages, e.g. to different cohorts of residents within the hospital, or all at once? The first paragraph stated that there is a 1 month observation phase followed by a 3 month intervention phase, but its not clear to me if this means the intervention continued over 3 months, or if the intervention was applied at the beginning of this phase, followed by ~3 months of patient follow-up. Further intervention in the study design and intervention sections could clarify this.

We have clarified that the training was conducted at the beginning of the three month intervention phase.

@@

> 2. Variables – more information on the injury severity score would be useful to readers, because I am unsure how this variable was planned to be used (it is not an outcome), and how precisely it was calculated. A reference is provided, but a high-level overview of the approach (including the range of potential scores) would help readers understand this measurement better.

We have added a description of the ISS to the methods section.

@@

> 3. “Differences in distributions of observed and extracted data” – I am unsure how observed data differs from extracted data – these terms were not previously explained or defined. Does it mean the observation period vs the 3 month intervention period? “distributions” is also somewhat unclear, histogram distributions, or a specific tendency statistic?

We have attempted to clarify what we mean by distributions (summary statistics) and directly observed data, in the section where we describe the feasibility outcomes.

@@

> 4. Statistical analysis: “We used an empty generalised linear mixed model to estimate the intracluster correlation coefficient” which outcomes does this apply to?

We have moved the description of this analysis to the supplementary materials, and clarified that it applies to all cause 30-day and in-hospital mortality.

@@

> 5. The analysis involved comparing the 1st month and 2nd-4th months, but it is not described how outcomes will be summed – e.g. consent rate, I imagine this the total consent rate in month 1 vs months 2-4? It would be helpful to directly confirm this in the text.

We have attempted to clarify that feasibility outcomes were not assessed as differences between the observation and intervention phases, but summarised across the entire period.

@@

> Results
> 6. Table 1. Missing counts are reported for some but not all measurements. Does the absence of a count indicate complete data? If so, a footnote stating this would help readers.

We have added a footnote to Table 1 to clarify that the absence of a count indicates complete data.

@@

> 7. Table 2. As mentioned in a prior comment it is unclear to me how these two groups are defined. Also, both have an N of 55, assuming these are patients it means there is a total N of 110, less than the total of 376. Why do the numbers differ?

We only extracted data from a convenience sample of patients to reduce the workload of the project officers collecting data. We have attempted to clarify this in the methods section and in the table caption.

@@

> 8. Table S1. Suggest clarifying that this is patient characteristics in the title.

We have added "patient" to the title.

> 9. Table S5. Certain cells are blank – a footnote should explain what this means (no observations?)

Yes, that is correct. Empty cells indicate that there were no observations. We have added a note to clarify this.

### 2025-07-07 manuscript.Rmd
Complete responding to comments

- Cluster randomized controlled trial comparing the effect of ATLS^®^, PTC and standard care on patient and provider outcomes.
- Prospective data collection with direct observations by dedicated research officers.
- Participating centers' heterogeneity may affect the study estimates and bias the results.
-->

@@

Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care is not possible. Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be more challenging.

@@

In each intervention arm the residents in one or two units were trained in either ATLS^®^ or PTC. For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit. We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients. The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai and the PTC training was conducted in New Delhi. Both trainings were conducted according to their respective standard curriculum [@acsAtls2018; @ptcfPtc2018], and we did not modify or adapt the delivery or content of these programmes during this pilot study.

The provider courses of both programmes are 2.5 days, and they teach how to assess, resuscitate and stabilize trauma patients and how to adapt to needs of different patient populations, for example pediatric, pregnant and geriatric patients. The teaching is based on case discussions and skill stations. There are some important differences between the two programs. The ATLS course include more on the inter-hospital transfer of patients and includes a greater emphasis on the trauma team [@acsAtls2018]. The PTC course on the other hand focuses on trauma care in the low resource setting [@ptcfPtc2018].

@@

@@

- Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all included patients.
- Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
- Differences in distributions between directly observed data and data extracted from medical records. This applied to all variables that could be reasonably expected to be present in the medical records, but we only extracted this data from a convenience sample of patients to reduce the workload of the project officers collecting data.
- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.

@@

@@

We wanted to include at least two clusters per arm to avoid drawing conclusions based on single centers, and we wanted to train at least two units per intervention cluster to be able to evaluate the logistics of sending residents for training. We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods. We anticipated that the numbers of patients included per cluster were anticipated would vary between clusters depending on the volume of patients at each hospital.<!--The time period was dictated by budget and time constraints.-->

@@

The research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained. All injuries were coded according to the International Classification of Diseases version 10 (ICD-10). Based on these ICD-10 codes, we calculated the Injury Severity Score using the R package `icdpicr` [@icdpicr_2022]. We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and health related quality of life (using the EQ-5D-3L). We did not calculate an EQ-5D-3L index score, because there is currently no Indian value set available for the EQ-5D-3L [@jyani_development_2022]. We also attempted to collect data on cause of death. A list of variables is available in Supplementary Table S2.<!--Supplemental Material 2 contains a full variable list, with definitions.-->

@@

```{r observed-vs-extracted, echo = FALSE, tab.cap = "Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview."}
observed.vs.extracted.comparison.table
```

@@

We show that it is feasible to conduct and collect data for a cluster randomized controlled trial comparing ATLS^®^ with PTC and standard care. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables, especially the cost of treatment (reported in Supplementary Table S2-13) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods. The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete. To improve completeness of data extracted from the medical records, a full scale trial should limit the number of variables extracted from the medical records and the importance of having these registered should be emphasised to the participating hospitals.

@@

All-cause 30-day mortality was missing in `r rate.lost.to.follow.up`% of patients. This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01% of patients [@Roberts2013; @Sierink2016]. Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow up patients after discharge. Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to previous trials. Following up patients after discharge is notoriously challenging in the setting of this study, and the full scale trial might need to focus on in-hospital mortality as the primary outcome. The difference in the number of patient deaths between the arms may have resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.

@@

During the course of this pilot study we deviated from the protocol in several ways. We have reported these deviations were appropriate in the methods section, and provide a detailed list as Supplementary
material S1. The most important deviation was the revision of the aim of this pilot study, because we initially planned to estimate potential effect sizes and other measures to help sample size calculations for a full scale trial, in addition to the feasibility outcomes. However, current guidance advice against using pilot studies to estimate effect sizes, as the usefulness of these estimates is questionable [INSERT REF]. We therefore decided to only report patient outcomes descriptively. Another important deviation from the protocol was the training of emergency medicine residents. We initially planned to train only surgical residents, but trauma management routines varied between participating hospitals, and we decided to adapt to local routines. The full scale study will need to accommodate this variation as part of the protocol.

There are some important limitations of this pilot study and therefore lessons to be learned and factored into the design of the full-scale trial. First, the volumes of patients included at some of the participating hospitals were lower than expected. A careful assessment of patient volumes as part of the screening process will be needed for the full-scale trial. Second, data on complications and cause of death were almost universally missing. To collect data on these variables will require other methods, as these were not explicitly described in the medical records and autopsy reports are not readily available. Third, we did not collect in-depth data on the standard care at each hospital. This data should be collected as part of the screening process for the full-scale trial. Fourth, we used sealed envelopes for randomisation, which increases the risk of bias and errors. The full scale trial should use a computer-generated randomisation system. Fifth, we did not blind the data analysts, but recommend doing so in the full scale trial. Finally, we assessed a very large number of potential outcomes, and the full scale trial should focus on the most relevant outcomes.

@@

Previous studies on the effect of ATLS^®^ or PTC training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016]. Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016]. In contrast, some studies have found that these programmes may be associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].

Considering the widespread use of trauma life support training, several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020]. Our study represents the first published attempt at a controlled trial of the effect of trauma life support training on patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot.

>>>>

- Cluster randomised controlled trial comparing the effect of ATLS^®^, PTC and standard care on patient and provider outcomes.
- Prospective data collection with direct observations by dedicated research officers.
- Participating centers' heterogeneity may affect the study estimates and bias the results.
-->

@@

Systematic reviews call for controlled trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], because conducting such effectiveness trials in settings where they are part of the standard of care is not possible. Many settings without routinely implemented trauma life support training are in low- and middle income countries, where trial logistics can be more challenging.

@@

@@

In each intervention arm the residents in one or two units were trained in either ATLS^®^ or PTC at the beginning of the three month intervention phase. For the purpose of this pilot study, our target was to train a minimum of 75% of residents in each unit. We did not train the units' faculty, because they are typically not directly involved in the initial management of trauma patients. The ATLS^®^ training was conducted in an ATLS^®^ certified training centre in Mumbai and the PTC training was conducted in New Delhi. Both trainings were conducted according to their respective standard curriculum [@acsAtls2018; @ptcfPtc2018], and we did not modify or adapt the delivery or content of these programmes during this pilot study.

The provider courses of both programmes are two days, and they teach how to assess, resuscitate and stabilize trauma patients and how to adapt to needs of different patient populations. The teaching is based on case discussions and skill stations. There are some important differences between the two programmes. The ATLS course include more on the inter-hospital transfer of patients and includes a greater emphasis on the trauma team [@acsAtls2018]. The PTC course on the other hand focuses on trauma care in the low resource setting [@ptcfPtc2018]. The ATLS^®^ programme is run by the American College of Surgeons and participants pay a fee to take a course, whereas the PTC programme is run by the UK Charity PTC Foundation and is free of charge.

@@

- Consent rate of patients and residents. This was equal to the percentage of patients or residents who consented to be included, out of the total number of eligible patients or residents. 
- Lost to follow up rate. This applied only to patients and was equal to the percentage of patients who did not complete 30 day follow up, out of all included patients.
- Missing data rate. This applied to each outcome and variable and was equal to the percentage of missing values.
- Differences in distributions between directly observed data and data extracted from medical records. Distribution refers to summary statistics and directly observed data refers to data collected by project officers while observing the delivery of care. This outcome applied to all variables that could be reasonably expected to be present in the medical records, but we decided to only extract this data from a convenience sample of patients to reduce the workload of the project officers collecting data.
- Pass rate. This applied only to residents in the intervention arms and was equal the percentage of residents who passed the training programme, out of the total number of trained residents.

We did not prespecify criteria to judge whether to continue with the full scale trial.

@@

We wanted to include at least two clusters per arm to avoid drawing conclusions based on single centers, and we wanted to train at least two units per intervention cluster to be able to evaluate the logistics of sending residents for training. We did not conduct a formal power calculation for this pilot study, as the purpose was to assess the feasibility of the trial logistics and research methods. We anticipated that the numbers of patients included per cluster were anticipated would vary between clusters depending on the volume of patients at each hospital.

@@

@@

The research officers collected data on demographics, vital signs, management details including imaging and surgery, and details of any injury sustained. All injuries were coded according to the International Classification of Diseases version 10 (ICD-10). Based on these ICD-10 codes, we calculated the Injury Severity Score (ISS) using the R package `icdpicr` [@icdpicr_2022]. The ISS is a widely used measure of injury severity and ranges from 0 to 75, with a cutoff of 16 often being used to define major trauma, and 75 representing unsurvivable trauma. We also collected data on potential outcomes for the full-scale trial, including mortality within 30-days and in-hospital, complications and health related quality of life (using the EQ-5D-3L). We did not calculate an EQ-5D-3L index score, because there is currently no Indian value set available for the EQ-5D-3L [@jyani_development_2022]. We also attempted to collect data on cause of death. A list of variables is available in Supplementary Table S2.<!--Supplemental Material 2 contains a full variable list, with definitions.-->

@@

```{r observed-vs-extracted, echo = FALSE, tab.cap = "Differences in distributions between directly observed data and data extracted from medical records, for selected variables that were collected through observation or interview in a convenience sample of patients."}
observed.vs.extracted.comparison.table
```

@@

We show that it is feasible to conduct and collect data for a cluster randomised controlled trial comparing ATLS^®^ with PTC and standard care. Missing data were low for key variables, including the primary outcome and many secondary outcomes. Some variables, especially the cost of treatment (reported in Supplementary Table S2-13) had very high missing data rates and may not be feasible to include in a full-scale trial, or require different data collection methods. The missing data was substantially higher when data was extracted from medical records instead of being directly observed, but the data were similarly distributed, indicating that data collected from medical records is reliable even if it is less complete. To improve completeness of data extracted from the medical records, a full scale trial should limit the number of variables extracted from the medical records and the importance of having these registered should be emphasised to the participating hospitals.

@@

All-cause 30-day mortality was missing in `r rate.lost.to.follow.up`% of patients. This may appear high, especially compared to for example the CRASH-2 and REACT-2 trials, which report missing primary outcome in less than 0.01% of patients [@Roberts2013; @Sierink2016]. Like many other trauma trials, both CRASH-2 and REACT-2 used in-hospital mortality as their primary outcome measure, whereas we attempted to follow up patients after discharge. Our missing data rate for in-hospital mortality was only `r p.missing.in.hospital.mortality`%, which is comparable to previous trials. Following up patients after discharge is notoriously challenging in the setting of this study, and the full scale trial might need to focus on in-hospital mortality as the primary outcome.

@@

During the course of this pilot study we deviated from the protocol in several ways. We have reported these deviations where appropriate in the methods section, and provide a list as Supplementary
material S1. The most important deviation was the revision of the aim of this pilot study, because we initially planned to estimate potential effect sizes and other measures to help sample size calculations for a full scale trial, in addition to the feasibility outcomes. However, current guidance advice against using pilot studies to estimate effect sizes, as the usefulness of these estimates is questionable [INSERT REF]. We therefore decided to only report patient outcomes descriptively. Another important deviation from the protocol was the training of emergency medicine residents. We initially planned to train only surgical residents, but trauma management routines varied between participating hospitals, and we decided to adapt to local routines. The full scale study will need to accommodate this variation as part of the protocol.

There are some important limitations of this pilot study and therefore additional lessons to be learned and factored into the design of the full-scale trial. First, the volumes of patients included at some of the participating hospitals were lower than expected. A careful assessment of patient volumes as part of the screening process will be needed for the full-scale trial. Second, data on complications and cause of death were almost universally missing. To collect data on these variables will require other methods, as these were not explicitly described in the medical records and autopsy reports are not readily available. Third, we did not collect in-depth data on the standard care at each hospital. This data should be collected as part of the screening process for the full-scale trial. Fourth, we used sealed envelopes for randomisation, which increases the risk of bias and errors. The full scale trial should use a computer-generated randomisation system. Fifth, we did not blind the data analysts, but recommend doing so in the full scale trial. Finally, we assessed a very large number of potential outcomes, and the full scale trial should focus on the most relevant outcomes.

@@

Previous studies on the effect of ATLS^®^ or PTC training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016]. Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016]. In contrast, some studies have found that these programmes may be associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016]. We found fewer deaths in the intervention arms compared to the standard arm. This difference  may have resulted from the randomisation process with a small number of heterogeneous clusters, highlighting the importance of taking varying cluster sizes into account when designing the full scale trial. A full scale trial remains ethical after this pilot study, considering that it was never powered to detect meaningful differences in clinical outcomes.

Considering the widespread use of trauma life support training, several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020]. Our study represents the first published attempt to pilot a controlled trial of the effect of trauma life support training on patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot study.

### 2025-07-07 supplementary-materials.Rmd
Complete responding to comments

We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient. We compared patients outcomes in all possible combinations of trial arms. In each combination we compared both differences in final values and differences in change from baseline. For the intervention arms the change from baseline was calculated as the difference between the one month period of data collection before the training was undertaken and the three month period after the training. For the control arm the data collection period was four months and the difference from baseline was calculated as the difference between the first one month and the following three months.

@@

# Table S2. Extended sample characteristics

>>>>

We used an empty generalised linear mixed model <!-- see [@Sommet2017] --> to estimate the intracluster correlation coefficient for all cause 30-day and in-hospital mortality. We compared patients outcomes in all possible combinations of trial arms. In each combination we compared both differences in final values and differences in change from baseline. For the intervention arms the change from baseline was calculated as the difference between the one month period of data collection before the training was undertaken and the three month period after the training. For the control arm the data collection period was four months and the difference from baseline was calculated as the difference between the first one month and the following three months.

@@

# Table S2. Extended patient sample characteristics
